A Wait-free Queue with Polylogarithmic Step Complexity

### 2023-01-10 podc_paper/algorithm.tex
edits to save space

    \If{\var{v} is a leaf node}
        \State\Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \Comment{required enqueue is in \var{v}.\fld{left}}
            \State \fld{dir} \assign\ \fld{left}
        \Else
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
        \EndIf \label{endChooseDir}
        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t. $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} \label{getRecurse}
    \EndIf
\EndFunction{GetEnqueue}
\end{algorithmic}
\caption{\label{pseudocode2}Queue implementation, continued.}
%\here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?  Not much; I tried it.}
\end{figure}

@@

%\paragraph{\tt{Refresh()} and \tt{Advance()}}
The goal of a \opemph{Refresh} on node \var{v} is to create a block representing the new blocks
in \var{v}'s
children and append it to \var{v}.\fld{blocks}. 
Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}
and Line \ref{invokeCreateBlock} creates the new block to be installed in \var{v}.\fld{blocks}[\var{h}].
If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, it means that there were no new blocks in \var{v}'s children to propagate to \var{v}.
In this case, \op{Refresh} returns true at line \ref{addOP} and terminates.
Otherwise, the CAS at line \ref{cas} attempts to install its new block into \var{v}.\fld{blocks}[\var{h}].
Either this CAS succeeds or some other process has installed a new block in this location.
Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head pointer 
from \var{h} to $\var{h}+1$
and to fill in the \fld{super} field of the most recently appended block.
The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
Since a \op{Refresh} may do a successful CAS but pause before calling \op{Advance} at line \ref{advance},
other processes help keep \fld{head} up to date by also calling \op{Advance}, 
either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on \var{v}.

\opemph{CreateBlock}(\var{v, i}) is used
by \op{Refresh} to construct a block to be installed in \var{v}.\fld{blocks}[\var{i}].
The block \nf{new} is created in Line \ref{initNewBlock}. 
The \eleft\ and \eright\ fields are filled in with the indices of the last blocks appended to \var{v}'s
children, which are obtained by reading the \fld{head} index in \var{v}'s children.
Recall that the \fld{sum\sub{enq}} field should store the number of enqueues in
\var{v}.\fld{blocks}[1..\var{i}].
Since these enqueues come from \var{v}.\fld{left.blocks}[1..\var{new}.\eleft] and \var{v}.\fld{blocks}[1..\var{new}.\eright],
the total number of these enqueues is $\var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right}.\fld{blocks}[\var{new}.\eright].\fld{sum\sub{enq}}$.
Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
subtracting from \var{new}.\fld{sum\sub{enq}} the number of enqueues that are already in \var{v}.\fld{blocks}[$1..\var{i}-1$].
The values of \var{new}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
Then, if \var{new}
is going to be installed in the root, line \ref{computeLength} computes the \fld{size} field, which
represents the number of elements in the queue after the operations in the block are performed.
Finally, if the newly created block contains no operations, \op{CreateBlock} returns \nl\ to indicate
that there is no need to append it to \var{v}.\fld{blocks}.

>>>>

    \If{\var{v} is a leaf node} \Return \var{v}.\fld{blocks}[\var{b}].\fld{element} \label{getBaseCase}
    \Else 
        \State \var{sum\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[\var{b}].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}$] from left child}
        \State \var{prev\sub{left}} \assign\ \var{v}.\fld{left.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eleft].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from left child}
        \State \var{prev\sub{right}} \assign\ \var{v}.\fld{right.blocks}[\var{v}.\fld{blocks}[$\var{b}-1$].\eright].\fld{sum\sub{enq}} \Comment{\#\ of enqueues in \var{v}.\fld{blocks}[1..$\var{b}-1$] from right child}
        \If{$\var{i} \leq \var{sum\sub{left}} - \var{prev\sub{left}}$} \label{leftOrRight} \Comment{required enqueue is in \var{v}.\fld{left}}
            \State \fld{dir} \assign\ \fld{left}
        \Else
            \State \fld{dir} \assign\ \fld{right}
            \State $\var{i}\ \assign\ \var{i} - (\var{sum\sub{left}} - \var{prev\sub{left}})$
        \EndIf \label{endChooseDir}
        \State \linecomment{Use binary search to find enqueue's block in \var{v}.\fld{dir} and its rank within block}
        \State find minimum $\var{b}'$ in range [\var{v}.\fld{blocks}[\var{b}-1].\edir+1..\var{v}.\fld{blocks}[\var{b}].\edir] s.t. $\var{v}.\fld{dir.blocks}[\var{b}'].\fld{sum\sub{enq}} \geq \var{i} + \var{prev\sub{dir}}$\label{getChild}
        \State $\var{i}'$ \assign\ $\var{i} - (\var{v}.\fld{dir.blocks}[\var{b}'-1].\fld{sum\sub{enq}} - \var{prev\sub{dir}})$\label{getChildIndex}
        \State \Return\Call{GetEnqueue}{\var{v}.\fld{dir}, $\var{b}'$, $\var{i}'$} \label{getRecurse}
    \EndIf
\EndFunction{GetEnqueue}
\end{algorithmic}
\caption{\label{pseudocode2}Queue implementation, continued.}
%\here{Would this routine be simpler to write if i parameter was the rank within whole node rather than within the block?  Not much; I tried it.}
\end{figure}

@@

%\paragraph{\tt{Refresh()} and \tt{Advance()}}
A \opemph{Refresh} on node \var{v} creates a block representing the new blocks
in \var{v}'s
children and tries to append it to \var{v}.\fld{blocks}. 
Line \ref{readHead} reads \var{v}.\fld{head} into the local variable \var{h}.
Line \ref{invokeCreateBlock} creates the new block to install in \var{v}.\fld{blocks}[\var{h}].
If line \ref{invokeCreateBlock} returns \nl\ instead of a new block, there were no new blocks in \var{v}'s children to propagate to \var{v}, so \op{Refresh} can return true at line \ref{addOP} and terminate.
Otherwise, the CAS at line \ref{cas} tries to install the new block into \var{v}.\fld{blocks}[\var{h}].
Either this CAS succeeds or some other process has installed a  block in this location.
Either way, line \ref{advance} then calls \opemph{Advance} to advance \var{v}'s head index 
from \var{h} to $\var{h}+1$
and fill in the \fld{super} field of the most recently appended block.
The boolean value returned by \op{Refresh} indicates whether its CAS succeeded.
A \op{Refresh} may pause after a successful CAS before calling \op{Advance} at line \ref{advance},
so other processes help keep \fld{head} up to date by  calling \op{Advance}, 
either at line \ref{helpAdvance} during a \op{Refresh} on \var{v}'s parent or line \ref{advance} during a \op{Refresh} on~\var{v}.

\opemph{CreateBlock}(\var{v, i}) is used
by \op{Refresh} to construct a block to be installed in \var{v}.\fld{blocks}[\var{i}].
The block \var{new} is created in Line \ref{initNewBlock}. 
The \eleft\ and \eright\ fields are filled in with the indices of the last blocks appended to \var{v}'s
children, which are obtained by reading the \fld{head} index in \var{v}'s children.
Since the \fld{sum\sub{enq}} field should store the number of enqueues in
\var{v}.\fld{blocks}[1..\var{i}] and these enqueues come from \var{v}.\fld{left.blocks}[1..\var{new}.\eleft] and \var{v}.\fld{blocks}[1..\var{new}.\eright], line \ref{createSumEnq} sets
\fld{sum\sub{enq}} to $\var{v}.\fld{left.blocks}[\var{new}.\eleft].\fld{sum\sub{enq}} + \var{v}.\fld{right}.\fld{blocks}[\var{new}.\eright].\fld{sum\sub{enq}}$.
Line \ref{computeNumEnq} sets \var{num\sub{enq}} to the number of enqueues in the new block by
subtracting from \var{new}.\fld{sum\sub{enq}} the number of enqueues that are already in \var{v}.\fld{blocks}[$1..\var{i}-1$].
The values of \var{new}.\fld{sum\sub{deq}} and \var{num\sub{deq}} are computed similarly.
Then, if \var{new}
is going to be installed in the root, line \ref{computeLength} computes the \fld{size} field, which
represents the number of elements in the queue after the operations in the block are performed.
Finally, if the new block contains no operations, \op{CreateBlock} returns \nl\ to indicate
 there is no need to install it.
